Formalized Linear Programming

### 2019-09-27 TeX/paper.tex
Move comments out to the border

@@

@@

@@

\section{Introduction}
\color{red}Describe motivation of project.
The aim is to provide a practically computable implementation of the simplex algorithm (Coq version did not aim to do this). It differs in that we use a more complex tableau data type (dead columns and unrestricted rows). We also use a different pivot rule (Bland's Rule - lexicographic rule seems to be an ambiguous name).

Description of why Lean is a good language to do this in.\color{black}

\section{Linear Programming}
\color{red} The standard simplex algorithm. No reference to Lean code \color{black}

The simplex algorithm aims to optimize the value of an objective function within the constraints of a set of affine inequalities. It can be used to solve problems of the following form.
% \color{red} This is a slightly nonstandard way of writing a linear program. The usual way would be $Ax \le b$. However the matrix that is stored during a simplex is $-A$ according to that representation, but $A$ according to the representation below.\color{black}
\begin{equation}
  \begin{aligned}
    \text{Find } x \text{ that maximizes } && c^Tx \\
    \text{subject to } && A x + b \ge 0\\
    \text{and } && x \ge 0
  \end{aligned}
\end{equation}
Here, $x \in \mathbb{Q}^n$, $c \in \mathbb{Q}^n$, $b \in \mathbb{Q}^m$, and $A \in \mathbb{Q}^{m \times n}$.
This can be written in an equivalent form \color{red} The inclusion of $z$ in the objective function is slightly unusual because obviously maximizing $c^Tx$ is the same as maximizing $c^Tx + z$. However $z$ is a cell in the tableau, it is the objective function value of the sample solution, and the explanation of the tableau is slightly easier with it\color{black}
\begin{equation}
  \begin{aligned}
    \text{Find } x_r \text{ and } x_c \text{ that maximize } && c^Tx_c + z \\
    \text{subject to } && x_r = A x_c + b \\
    \text{and } && x_r \ge 0, x_c \ge 0
  \end{aligned}
\end{equation}
with $x_r \in \mathbb{Q}^m$, $x_c \in \mathbb{Q}^n$ and $z \in \mathbb{Q}$

@@

The simplex algorithm iteratively performs the \textit{pivot} operation on a tableau. The pivot operation takes a row and a column index as an argument, and swaps their positions in the tableau. If $x_{r_i}$ and $x_{c_j}$ are pivoted, then in the new tableau $x_{r_i}$ becomes a row variable and $x_{c_j}$ becomes a column variable. The variables $(x_{r_1}, x_{r_2}, \dots, x_{c_j}, \dots x_{r_m})$ are expressed as affine functions of $(x_{c_1}, x_{c_2}, \dots, x_{r_i}, \dots, x_{c_n})$. The matrix $A$, and the vectors $c$ and $b$ are updated to $A'$, $c'$ and $b'$, such that the tableau represents the same subset of $\mathbb{Q}^{m + n}$, and the same optimization problem. Additionally the vectors $r$ and $c$ are updated to vectors $r' := (r_1, r_2, \dots c_j, \dots, r_m)$ and $c' := (c_1, c_2 \dots r_i, \dots, c_n)$. \color{red} Oops, two things are called $c$ \color{black} Preserving the same polyhedron is possible if $A_{ij} \ne 0$.
\begin{equation}
  \begin{aligned}
    x_{r_1} && = && A'_{11}x_{c_1} && + && A'_{12}x_{c_2} && + && \dots && + && A'_{1j}x_{r_i} && + && \dots && + && A'_{1n}x_{c_n} && + && b'_1\\
    x_{r_2} && = && A'_{21}x_{c_1} && + && A'_{22}x_{c_2} && + && \dots && + && A'_{2j}x_{r_i} && + && \dots && + && A'_{2n}x_{c_n} && + && b'_2\\
    \vdots && && \vdots && && \vdots && && && && \vdots && && && && \vdots && && \vdots \\
    x_{c_j} && = && A'_{i1}x_{c_1} && + && A'_{i2}x_{c_2} && + && \dots && + && A'_{ij}x_{r_i}&& + && \dots && + && A'_{in}x_{c_n} && + && b'_i\\
    \vdots && && \vdots && && \vdots && && && && \vdots && && && && \vdots && && \vdots \\
    x_{r_m} && = && A'_{m1}x_{c_1} && + && A'_{m2}x_{c_2} && + && \dots && + && A'_{mj}x_{r_i}&& + && \dots && + && A'_{mn}x_{c_n} && + && b'_n\\
    c^Tx + z && = && c'_1x_{c_1} && + && c'_2x_{c_2} && + && \dots && + && c'_jx_{r_i} && + && \dots && + && c'_nx_{c_n} && + && z'
  \end{aligned}
\end{equation}
The expressions for the updated matrix $A'$, and the updated vectors $c'$, $b'$ are given by the following formulas \cite{Nelson:1980:TPV:909447}.
\begin{equation}\label{eqn:pivot_tableau}
A'_{i'j'} = \begin{cases}
  A_{ij}^{-1} & \text{if } i' = i \text{ and } j' = j \\
  -A_{ij'}/A_{ij} & \text{if } i' = i \text{ and } j' \ne j \\
  A_{i'j}/A_{ij} & \text{if } i' \ne i \text{ and } j' = j \\
  A_{i'j'} - A_{i'j}A_{ij'}/A_{ij}  & \text{if } i' \ne i \text{ and } j' \ne j
\end{cases}
\end{equation}
\begin{equation}\label{eqn:pivot_const}
  b'_{i'} = \begin{cases}
    -b_i/A_{ij} & \text{if } i' = i \\
    b_{i'} - A_{i'j}b_i/A_{ij} & \text{if } i' \ne i
  \end{cases}
\end{equation}
\begin{equation}
  c'_{j'} = \begin{cases}
    c_j/A_{ij} & \text{if } j' = j \\
    c_{j'} - c_jA_{ij'}/A_{ij} & \text{if } j' \ne j
  \end{cases}
\end{equation}
\begin{equation}\label{eqn:pivot_obj}
  z' = z - c_jb_i/A_{ij}
\end{equation}

Note that $b$, $c$ and $z$ are updated as though they are a continuation of $A$; $b$ is updated in the same way as a column of $A$, and $c$ is updated in the same way as a row of $A$. \color{red} This is quite important in the understanding of our version of the simplex. In our version we just optimize a variable $x_i$, rather than having an objective function, but this is actually the exact same thing.\color{black}

@@

\color{red} Explanation below could be better\color{black}

@@

\section{Description of our version of the linear programming problem}
\color{red} Our version, dead columns, unrestricted variables, stopping early. Also very little reference to Lean code. \color{black}

@@

\section{Lean implementation}
\color{red} Only describe Lean code where I had to do something in a way that was not the obvious way. The termination proof is interesting since the majority of the length of the proof is about stuff that was not mentioned on paper. Also `pequiv` made a few things a little easier and is not obvious. Also state correctness theorems \color{black}

@@

\color{red} The full correctness statement is actually quite complicated, but there are simpler versions that state correctness of the most important parts of the algorithm (it returns the correct one out of optimal unbounded or while)\color{black}
If the simplex algorithm is correct, then the tableau returned should be a feasible tableau, representing the same polyhedron as the starting tableau. Most importantly we also need to prove that it returns the correct condition out of \lstinline|while|, \lstinline|unbounded|, and \lstinline|optimal|.

@@

\subsection{pequiv}
\color{red} I can't really claim confidently that this was any better than any other way of doing this. My initial instinct when starting the project was that proving properties of matrices defined using \lstinline|if ... then ... else ...| would be extremely difficult, however the pivot defined like this was easier to use, and I think the proofs mentioned in this section would be shorter overall with \lstinline| if ... then ... else ...| as well, because of the lines spent proving the properties of the matrices $R$ and $C$ mentioned below. \color{black}
The simplex algorithm requires a lot of reasoning about minors of matrices or vectors. Taking a minor of a matrix is a linear map, and can be expressed as a matrix multiplication. Within the simplex algorithm there are two matrix minors that are often taken, the minor of a vector within a polyhedron corresponding to the column variables, and the minor for the row variables.

@@

\color{red} Do I want to go into as much detail as below\color{black}

@@

\color{red}
How much detail to go into in this proof? This was one of the more challenging parts of the formalization, however, the proof is actually not very nicely written at the moment, for example the definition of fickle is not quite the same as in Chvatal. I used a weaker notion that seemed good enough,but then one variable that should be fickle ended up not being fickle, and I had to get round this in a slightly messy way.
\color{black}

@@

\section{Refined algorithm}
\color{red} Say that we refined it and it is vaguely fast. Do I go into detail about how we did it. Probably quite a short section. \color{black}

The simplex algorithm was originally written on an extremely slow implementation of matrices. The implementation in the Lean maths library \color{red}[cite mathlib]\color{black} is as the type of binary functions from two finite sets into a ring, in our case $\mathbb{Q}$. This implementation is convenient for writing proofs, but very slow to compute. In order to have a practical executable version of the simplex algorithm, it was necessary to transfer the algorithm and proofs onto a faster implementation of matrices using arrays. Ideally this would be done using a refinement framework \color{red} brief explanation of what that means here. Would it be possible to generalize this proof onto a type class rather than a concrete type using this framework?\color{black}. This is not currently available in Lean, so it was instead done in an ad hoc way, using a tableau type class. The proof of the simplex algorithm can be generalized onto any type for which there is an instance of the \lstinline|is_tableau| type class.

@@

By using the \lstinline|is_tableau| type class, implementing and proving correctness of a fast implementation of the simplex algorithm, is reduced to defining an instance of this class. To define an instance of this class, it is necessary to provide a function \lstinline|to_tableau| from a fast tableau type into \lstinline|tableau m n|, and three functions \lstinline|pivot_col|, \lstinline|pivot_row|, and \lstinline|pivot|. The only proofs about these functions that must be provided are that they agree with the functions on \lstinline|tableau m n|. \color{red} This sentence needs to be changed \color{black}

\section{Conclusion}
\color{red}
Follow up work. Difficulties I came across. It would have been nice to have a generic theory of linear programming. Some of what I wrote could be applied to other implementations of simplex, but not much.
\color{black}
\bibliography{bibl}

>>>>

\def\paperversiondraft{draft}
\def\paperversionblind{blind}

\ifx\paperversion\paperversionblind
\else
  \def\paperversion{draft}
\fi

@@

\usepackage{xargs}
\usepackage{todonotes}
\usepackage{xparse}
\usepackage{xifthen, xstring}
\usepackage{ulem}
\usepackage{xspace}
\makeatletter
\font\uwavefont=lasyb10 scaled 652

\newcommand\colorwave[1][blue]{\bgroup\markoverwith{\lower3\p@\hbox{\uwavefont\textcolor{#1}{\char58}}}\ULon}
\makeatother

\ifx\paperversion\paperversiondraft
\newcommand\createtodoauthor[2]{%
\def\tmpdefault{emptystring}
\expandafter\newcommand\csname #1\endcsname[2][\tmpdefault]{\def\tmp{##1}\ifthenelse{\equal{\tmp}{\tmpdefault}}
   {\todo[linecolor=#2!20,backgroundcolor=#2!25,bordercolor=#2,size=\tiny]{\textbf{#1:} ##2}}
   {\ifthenelse{\equal{##2}{}}{\colorwave[#2]{##1}\xspace}{\todo[linecolor=#2!10,backgroundcolor=#2!25,bordercolor=#2]{\tiny \textbf{#1:} ##2}\colorwave[#2]{##1}}}}}
\else
\newcommand\createtodoauthor[2]{%
\expandafter\newcommand\csname #1\endcsname[2][\@nil]{}}
\fi

\createtodoauthor{chris}{red}
\createtodoauthor{tobias}{blue}
\createtodoauthor{authorThree}{green}
\createtodoauthor{authorFour}{orange}
\createtodoauthor{authorFive}{purple}

@@

% Broaden margins to make room for todo notes
\ifx\paperversion\paperversiondraft
  \paperwidth=\dimexpr \paperwidth + 6cm\relax
  \oddsidemargin=\dimexpr\oddsidemargin + 3cm\relax
  \evensidemargin=\dimexpr\evensidemargin + 3cm\relax
  \marginparwidth=\dimexpr \marginparwidth + 3cm\relax
  \setlength{\marginparwidth}{4cm}
\fi

@@

\section{Introduction}
\chris{
Describe motivation of project.
The aim is to provide a practically computable implementation of the simplex algorithm (Coq version did not aim to do this). It differs in that we use a more complex tableau data type (dead columns and unrestricted rows). We also use a different pivot rule (Bland's Rule - lexicographic rule seems to be an ambiguous name).

Description of why Lean is a good language to do this in.}

\section{Linear Programming}
\chris{The standard simplex algorithm. No reference to Lean code}

The simplex algorithm aims to optimize the value of an objective function within the constraints of a set of affine inequalities. It can be used to solve problems of the following form.
% \color{red} This is a slightly nonstandard way of writing a linear program. The usual way would be $Ax \le b$. However the matrix that is stored during a simplex is $-A$ according to that representation, but $A$ according to the representation below.\color{black}
\begin{equation}
  \begin{aligned}
    \text{Find } x \text{ that maximizes } && c^Tx \\
    \text{subject to } && A x + b \ge 0\\
    \text{and } && x \ge 0
  \end{aligned}
\end{equation}
Here, $x \in \mathbb{Q}^n$, $c \in \mathbb{Q}^n$, $b \in \mathbb{Q}^m$, and $A \in \mathbb{Q}^{m \times n}$.
This can be written in an equivalent form \chris{The inclusion of $z$ in the objective function is slightly unusual because obviously maximizing $c^Tx$ is the same as maximizing $c^Tx + z$. However $z$ is a cell in the tableau, it is the objective function value of the sample solution, and the explanation of the tableau is slightly easier with it}
\begin{equation}
  \begin{aligned}
    \text{Find } x_r \text{ and } x_c \text{ that maximize } && c^Tx_c + z \\
    \text{subject to } && x_r = A x_c + b \\
    \text{and } && x_r \ge 0, x_c \ge 0
  \end{aligned}
\end{equation}
with $x_r \in \mathbb{Q}^m$, $x_c \in \mathbb{Q}^n$ and $z \in \mathbb{Q}$

@@

The simplex algorithm iteratively performs the \textit{pivot} operation on a tableau. The pivot operation takes a row and a column index as an argument, and swaps their positions in the tableau. If $x_{r_i}$ and $x_{c_j}$ are pivoted, then in the new tableau $x_{r_i}$ becomes a row variable and $x_{c_j}$ becomes a column variable. The variables $(x_{r_1}, x_{r_2}, \dots, x_{c_j}, \dots x_{r_m})$ are expressed as affine functions of $(x_{c_1}, x_{c_2}, \dots, x_{r_i}, \dots, x_{c_n})$. The matrix $A$, and the vectors $c$ and $b$ are updated to $A'$, $c'$ and $b'$, such that the tableau represents the same subset of $\mathbb{Q}^{m + n}$, and the same optimization problem. Additionally the vectors $r$ and $c$ are updated to vectors $r' := (r_1, r_2, \dots c_j, \dots, r_m)$ and $c' := (c_1, c_2 \dots r_i, \dots, c_n)$.
\chris{Oops, two things are called $c$} Preserving the same polyhedron is possible if $A_{ij} \ne 0$.
\begin{equation}
  \begin{aligned}
    x_{r_1} && = && A'_{11}x_{c_1} && + && A'_{12}x_{c_2} && + && \dots && + && A'_{1j}x_{r_i} && + && \dots && + && A'_{1n}x_{c_n} && + && b'_1\\
    x_{r_2} && = && A'_{21}x_{c_1} && + && A'_{22}x_{c_2} && + && \dots && + && A'_{2j}x_{r_i} && + && \dots && + && A'_{2n}x_{c_n} && + && b'_2\\
    \vdots && && \vdots && && \vdots && && && && \vdots && && && && \vdots && && \vdots \\
    x_{c_j} && = && A'_{i1}x_{c_1} && + && A'_{i2}x_{c_2} && + && \dots && + && A'_{ij}x_{r_i}&& + && \dots && + && A'_{in}x_{c_n} && + && b'_i\\
    \vdots && && \vdots && && \vdots && && && && \vdots && && && && \vdots && && \vdots \\
    x_{r_m} && = && A'_{m1}x_{c_1} && + && A'_{m2}x_{c_2} && + && \dots && + && A'_{mj}x_{r_i}&& + && \dots && + && A'_{mn}x_{c_n} && + && b'_n\\
    c^Tx + z && = && c'_1x_{c_1} && + && c'_2x_{c_2} && + && \dots && + && c'_jx_{r_i} && + && \dots && + && c'_nx_{c_n} && + && z'
  \end{aligned}
\end{equation}
The expressions for the updated matrix $A'$, and the updated vectors $c'$, $b'$ are given by the following formulas \cite{Nelson:1980:TPV:909447}.
\begin{equation}\label{eqn:pivot_tableau}
A'_{i'j'} = \begin{cases}
  A_{ij}^{-1} & \text{if } i' = i \text{ and } j' = j \\
  -A_{ij'}/A_{ij} & \text{if } i' = i \text{ and } j' \ne j \\
  A_{i'j}/A_{ij} & \text{if } i' \ne i \text{ and } j' = j \\
  A_{i'j'} - A_{i'j}A_{ij'}/A_{ij}  & \text{if } i' \ne i \text{ and } j' \ne j
\end{cases}
\end{equation}
\begin{equation}\label{eqn:pivot_const}
  b'_{i'} = \begin{cases}
    -b_i/A_{ij} & \text{if } i' = i \\
    b_{i'} - A_{i'j}b_i/A_{ij} & \text{if } i' \ne i
  \end{cases}
\end{equation}
\begin{equation}
  c'_{j'} = \begin{cases}
    c_j/A_{ij} & \text{if } j' = j \\
    c_{j'} - c_jA_{ij'}/A_{ij} & \text{if } j' \ne j
  \end{cases}
\end{equation}
\begin{equation}\label{eqn:pivot_obj}
  z' = z - c_jb_i/A_{ij}
\end{equation}

Note that $b$, $c$ and $z$ are updated as though they are a continuation of $A$; $b$ is updated in the same way as a column of $A$, and $c$ is updated in the same way as a row of $A$.
\chris{This is quite important in the understanding of our version of the simplex. In our version we just optimize a variable $x_i$, rather than having an objective function, but this is actually the exact same thing.}

@@

\chris{Explanation below could be better}

@@

\section{Description of our version of the linear programming problem}
\chris{Our version, dead columns, unrestricted variables, stopping early. Also very little reference to Lean code.}

@@

\section{Lean implementation}
\chris{Only describe Lean code where I had to do something in a way that was not the obvious way. The termination proof is interesting since the majority of the length of the proof is about stuff that was not mentioned on paper. Also `pequiv` made a few things a little easier and is not obvious. Also state correctness theorems}

@@

\chris{The full correctness statement is actually quite complicated, but there are simpler versions that state correctness of the most important parts of the algorithm (it returns the correct one out of optimal unbounded or while)}
If the simplex algorithm is correct, then the tableau returned should be a feasible tableau, representing the same polyhedron as the starting tableau. Most importantly we also need to prove that it returns the correct condition out of \lstinline|while|, \lstinline|unbounded|, and \lstinline|optimal|.

@@

\subsection{pequiv}
\chris{I can't really claim confidently that this was any better than any other way of doing this. My initial instinct when starting the project was that proving properties of matrices defined using \lstinline|if ... then ... else ...| would be extremely difficult, however the pivot defined like this was easier to use, and I think the proofs mentioned in this section would be shorter overall with \lstinline| if ... then ... else ...| as well, because of the lines spent proving the properties of the matrices $R$ and $C$ mentioned below.}
The simplex algorithm requires a lot of reasoning about minors of matrices or vectors. Taking a minor of a matrix is a linear map, and can be expressed as a matrix multiplication. Within the simplex algorithm there are two matrix minors that are often taken, the minor of a vector within a polyhedron corresponding to the column variables, and the minor for the row variables.

@@

\chris{Do I want to go into as much detail as below}

@@

\chris{How much detail to go into in this proof? This was one of the more challenging parts of the formalization, however, the proof is actually not very nicely written at the moment, for example the definition of fickle is not quite the same as in Chvatal. I used a weaker notion that seemed good enough,but then one variable that should be fickle ended up not being fickle, and I had to get round this in a slightly messy way.}

@@

\section{Refined algorithm}
\chris{Say that we refined it and it is vaguely fast. Do I go into detail about how we did it. Probably quite a short section.}

The simplex algorithm was originally written on an extremely slow implementation of matrices. The implementation in the Lean maths library \chris{cite mathlib} is as the type of binary functions from two finite sets into a ring, in our case $\mathbb{Q}$. This implementation is convenient for writing proofs, but very slow to compute. In order to have a practical executable version of the simplex algorithm, it was necessary to transfer the algorithm and proofs onto a faster implementation of matrices using arrays. Ideally this would be done using a refinement framework \chris{brief explanation of what that means here. Would it be possible to generalize this proof onto a type class rather than a concrete type using this framework?}. This is not currently available in Lean, so it was instead done in an ad hoc way, using a tableau type class. The proof of the simplex algorithm can be generalized onto any type for which there is an instance of the \lstinline|is_tableau| type class.

@@

By using the \lstinline|is_tableau| type class, implementing and proving correctness of a fast implementation of the simplex algorithm, is reduced to defining an instance of this class. To define an instance of this class, it is necessary to provide a function \lstinline|to_tableau| from a fast tableau type into \lstinline|tableau m n|, and three functions \lstinline|pivot_col|, \lstinline|pivot_row|, and \lstinline|pivot|. The only proofs about these functions that must be provided are that they agree with the functions on \lstinline|tableau m n|. \chris{This sentence needs to be changed}

\section{Conclusion}
\chris{Follow up work. Difficulties I came across. It would have been nice to have a generic theory of linear programming. Some of what I wrote could be applied to other implementations of simplex, but not much.
}
\bibliography{bibl}
